---
title: "Ch13_S3-3"
author: "Min-Yao"
date: "2023-03-12"
output: html_document
---

## 13.6 Inheritance {#s3-inheritance}
\index{S3!inheritance}
\index{S3!methods!inheriting}
\index{inheritance!S3}

S3 classes can share behaviour through a mechanism called __inheritance__. Inheritance is powered by three ideas:

*   The class can be a character _vector_. For example, the `ordered` and 
    `POSIXct` classes have two components in their class:

    ```{r}
    class(ordered("x"))
    class(Sys.time())
    ```
    \indexc{POSIXct}

*   If a method is not found for the class in the first element of the 
    vector, R looks for a method for the second class (and so on):
  
    ```{r}
    s3_dispatch(print(ordered("x")))
    s3_dispatch(print(Sys.time()))
    ```
  
*   A method can delegate work by calling `NextMethod()`. We'll come back to 
    that very shortly; for now, note that `s3_dispatch()` reports delegation 
    with `->`.
   
    ```{r}
    s3_dispatch(ordered("x")[1])
    s3_dispatch(Sys.time()[1])
    ```

Before we continue we need a bit of vocabulary to describe the relationship between the classes that appear together in a class vector. We'll say that `ordered` is a __subclass__ of `factor` because it always appears before it in the class vector, and, conversely, we'll say `factor` is a __superclass__ of `ordered`. 

S3 imposes no restrictions on the relationship between sub- and superclasses but your life will be easier if you impose some. I recommend that you adhere to two simple principles when creating a subclass:

* The base type of the subclass should be that same as the superclass.

* The attributes of the subclass should be a superset of the attributes
  of the superclass.

`POSIXt` does not adhere to these principles because `POSIXct` has type double, and `POSIXlt` has type list. This means that `POSIXt` is not a superclass, and illustrates that it's quite possible to use the S3 inheritance system to implement other styles of code sharing (here `POSIXt` plays a role more like an interface), but you'll need to figure out safe conventions yourself.
\indexc{POSIXt}

### 13.6.1 `NextMethod()`
\indexc{NextMethod()}

`NextMethod()` is the hardest part of inheritance to understand, so we'll start with a concrete example for the most common use case: `[`.  We'll start by creating a simple toy class: a `secret` class that hides its output when printed:

```{r}
new_secret <- function(x = double()) {
  stopifnot(is.double(x))
  structure(x, class = "secret")
}

print.secret <- function(x, ...) {
  print(strrep("x", nchar(x)))
  invisible(x)
}

x <- new_secret(c(15, 1, 456))
x
```

This works, but the default `[` method doesn't preserve the class:

```{r}
s3_dispatch(x[1])
x[1]
```

To fix this, we need to provide a `[.secret` method. How could we implement this method? The naive approach won't work because we'll get stuck in an infinite loop:

```{r}
`[.secret` <- function(x, i) {
  new_secret(x[i])
}
```

Instead, we need some way to call the underlying `[` code, i.e. the implementation that would get called if we didn't have a `[.secret` method. One approach would be to `unclass()` the object:

```{r}
`[.secret` <- function(x, i) {
  x <- unclass(x)
  new_secret(x[i])
}
x[1]
```

This works, but is inefficient because it creates a copy of `x`. A better approach is to use `NextMethod()`, which concisely solves the problem of delegating to the method that would have been called if `[.secret` didn't exist:

```{r}
`[.secret` <- function(x, i) {
  new_secret(NextMethod())
}
x[1]
```

We can see what's going on with `sloop::s3_dispatch()`:

```{r}
s3_dispatch(x[1])
```

The `=>` indicates that `[.secret` is called, but that `NextMethod()` delegates work to the underlying internal `[` method, as shown by the `->`.

As with `UseMethod()`, the precise semantics of `NextMethod()` are complex. In particular, it tracks the list of potential next methods with a special variable, which means that modifying the object that's being dispatched upon will have no impact on which method gets called next.

### 13.6.2 Allowing subclassing {#s3-subclassing}
\index{S3!subclassing}

When you create a class, you need to decide if you want to allow subclasses, because it requires some changes to the constructor and careful thought in your methods. 

To allow subclasses, the parent constructor needs to have `...` and `class` arguments:

```{r}
new_secret <- function(x, ..., class = character()) {
  stopifnot(is.double(x))

  structure(
    x,
    ...,
    class = c(class, "secret")
  )
}
```

Then the subclass constructor can just call to the parent class constructor with additional arguments as needed. For example, imagine we want to create a supersecret class which also hides the number of characters:

```{r}
new_supersecret <- function(x) {
  new_secret(x, class = "supersecret")
}

print.supersecret <- function(x, ...) {
  print(rep("xxxxx", length(x)))
  invisible(x)
}

x2 <- new_supersecret(c(15, 1, 456))
x2
```

To allow inheritance, you also need to think carefully about your methods, as you can no longer use the constructor. If you do, the method will always return the same class, regardless of the input. This forces whoever makes a subclass to do a lot of extra work. 

Concretely, this means we need to revise the `[.secret` method. Currently it always returns a `secret()`, even when given a supersecret:

```{r}
`[.secret` <- function(x, ...) {
  new_secret(NextMethod())
}

x2[1:3]
```

\indexc{vec\_restore()}
We want to make sure that `[.secret` returns the same class as `x` even if it's a subclass. As far as I can tell, there is no way to solve this problem using base R alone. Instead, you'll need to use the vctrs package, which provides a solution in the form of the `vctrs::vec_restore()` generic. This generic takes two inputs: an object which has lost subclass information, and a template object to use for restoration.

Typically `vec_restore()` methods are quite simple: you just call the constructor with appropriate arguments:

```{r}
vec_restore.secret <- function(x, to, ...) new_secret(x)
vec_restore.supersecret <- function(x, to, ...) new_supersecret(x)
```

(If your class has attributes, you'll need to pass them from `to` into the constructor.)

Now we can use `vec_restore()` in the `[.secret` method:

```{r}
`[.secret` <- function(x, ...) {
  vctrs::vec_restore(NextMethod(), x)
}
x2[1:3]
```

(I only fully understood this issue quite recently, so at time of writing it is not used in the tidyverse. Hopefully by the time you're reading this, it will have rolled out, making it much easier to (e.g.) subclass tibbles.)

If you build your class using the tools provided by the vctrs package, `[` will gain this behaviour automatically. You will only need to provide your own `[` method if you use attributes that depend on the data or want non-standard subsetting behaviour. See `?vctrs::new_vctr` for details.

### 13.6.3 Exercises

1.  How does `[.Date` support subclasses? How does it fail to support 
    subclasses?

2.  R has two classes for representing date time data, `POSIXct` and
    `POSIXlt`, which both inherit from `POSIXt`. Which generics have
    different behaviours for the two classes? Which generics share the same
    behaviour?

3.  What do you expect this code to return? What does it actually return?
    Why?

    ```{r, eval = FALSE}
    generic2 <- function(x) UseMethod("generic2")
    generic2.a1 <- function(x) "a1"
    generic2.a2 <- function(x) "a2"
    generic2.b <- function(x) {
      class(x) <- "a1"
      NextMethod()
    }

    generic2(structure(list(), class = c("b", "a2")))
    ```

## 13.7 Dispatch details {#s3-dispatch}
\index{S3!method dispatch}

This chapter concludes with a few additional details about method dispatch. It is safe to skip these details if you're new to S3.

### 13.7.1 S3 and base types {#implicit-class}
\index{implicit class}
\index{S3!implicit class}

What happens when you call an S3 generic with a base object, i.e. an object with no class? You might think it would dispatch on what `class()` returns:

```{r}
class(matrix(1:5))
```

But unfortunately dispatch actually occurs on the __implicit class__, which has three components:

* The string "array" or "matrix" if the object has dimensions
* The result of `typeof()` with a few minor tweaks
* The string "numeric" if object is "integer" or "double"

There is no base function that will compute the implicit class, but you can use `sloop::s3_class()`

```{r}
s3_class(matrix(1:5))
```

This is used by `s3_dispatch()`:

```{r}
s3_dispatch(print(matrix(1:5)))
```

This means that the `class()` of an object does not uniquely determine its dispatch:

```{r}
x1 <- 1:5
class(x1)
s3_dispatch(mean(x1))

x2 <- structure(x1, class = "integer")
class(x2)
s3_dispatch(mean(x2))
```

### 13.7.2 Internal generics {#internal-generics}
\index{generics!internal}

Some base functions, like `[`, `sum()`, and `cbind()`, are called __internal generics__ because they don't call `UseMethod()` but instead  call the C functions `DispatchGroup()` or `DispatchOrEval()`. `s3_dispatch()` shows internal generics by including the name of the generic followed by `(internal)`:

```{r}
s3_dispatch(Sys.time()[1])
```

For performance reasons, internal generics do not dispatch to methods unless the class attribute has been set, which means that internal generics do not use the implicit class. Again, if you're ever confused about method dispatch, you can rely on `s3_dispatch()`.

### 13.7.3 Group generics
\index{S3!group generics}
\index{generics!group}

Group generics are the most complicated part of S3 method dispatch because they involve both `NextMethod()` and internal generics. Like internal generics, they only exist in base R, and you cannot define your own group generic.

There are four group generics: 

* __Math__: `abs()`, `sign()`, `sqrt()`, `floor()`, `cos()`, `sin()`, `log()`,
  and more (see `?Math` for the complete list).

* __Ops__: `+`, `-`, `*`, `/`, `^`, `%%`, `%/%`, `&`, `|`, `!`, `==`, `!=`, `<`,
  `<=`, `>=`, and `>`.

* __Summary__: `all()`, `any()`, `sum()`, `prod()`, `min()`, `max()`, and 
  `range()`.

* __Complex__: `Arg()`, `Conj()`, `Im()`, `Mod()`, `Re()`.

Defining a single group generic for your class overrides the default behaviour for all of the members of the group. Methods for group generics are looked for only if the methods for the specific generic do not exist:

```{r}
s3_dispatch(sum(Sys.time()))
```

Most group generics involve a call to `NextMethod()`. For example, take `difftime()` objects. If you look at the method dispatch for `abs()`, you'll see there's a `Math` group generic defined.

```{r}
y <- as.difftime(10, units = "mins")
s3_dispatch(abs(y))
```

`Math.difftime` basically looks like this:

```{r}
Math.difftime <- function(x, ...) {
  new_difftime(NextMethod(), units = attr(x, "units"))
}
```

It dispatches to the next method, here the internal default, to perform the actual computation, then restore the class and attributes. (To better support subclasses of `difftime` this would need to call `vec_restore()`, as described in Section \@ref(s3-subclassing).)

Inside a group generic function a special variable `.Generic` provides the actual generic function called. This can be useful when producing error messages, and can sometimes be useful if you need to manually re-call the generic with different arguments.

### 13.7.4 Double dispatch
\index{double dispatch}
\index{method dispatch!S3!double dispatch}

Generics in the Ops group, which includes the two-argument arithmetic and Boolean operators like `-` and `&`, implement a special type of method dispatch. They dispatch on the type of _both_ of the arguments, which is called __double dispatch__. This is necessary to preserve the commutative property of many operators, i.e. `a + b` should equal `b + a`. Take the following simple example:

```{r}
date <- as.Date("2017-01-01")
integer <- 1L

date + integer
integer + date
```

If `+` dispatched only on the first argument, it would return different values for the two cases. To overcome this problem, generics in the Ops group use a slightly different strategy from usual. Rather than doing a single method dispatch, they do two, one for each input. There are three possible outcomes of this lookup:

* The methods are the same, so it doesn't matter which method is used.

* The methods are different, and R falls back to the internal method with 
  a warning.

* One method is internal, in which case R calls the other method.

This approach is error prone so if you want to implement robust double dispatch for algebraic operators, I recommend using the vctrs package. See `?vctrs::vec_arith` for details.

### 13.7.5 Exercises

1.  Explain the differences in dispatch below:

    ```{r}
    length.integer <- function(x) 10
    
    x1 <- 1:5
    class(x1)
    s3_dispatch(length(x1))
    
    x2 <- structure(x1, class = "integer")
    class(x2)
    s3_dispatch(length(x2))
    ```

2.  What classes have a method for the `Math` group generic in base R? Read
    the source code. How do the methods work?

3.  `Math.difftime()` is more complicated than I described. Why?