---
title: "Ch13_S3-2"
author: "Min-Yao"
date: "2023-03-05"
output: 
  html_document: 
    keep_md: yes
---

```{r}
library(sloop)
library(tidyverse)
library(purrr)
library(pryr)
```


## 13.4 Generics and methods {#s3-methods}
\indexc{UseMethod()} 
\index{S3!generics}
\index{generics!S3}

The job of an S3 generic is to perform method dispatch, i.e. find the specific implementation for a class. Method dispatch is performed by `UseMethod()`, which every generic calls[^internal-generic]. `UseMethod()` takes two arguments: the name of the generic function (required), and the argument to use for method dispatch (optional). If you omit the second argument, it will dispatch based on the first argument, which is almost always what is desired.

[^internal-generic]: The exception is internal generics, which are implemented in C, and are the topic of Section \@ref(internal-generics).

Most generics are very simple, and consist of only a call to `UseMethod()`. Take `mean()` for example:

```{r}
mean
```

Creating your own generic is similarly simple:

```{r}
my_new_generic <- function(x) {
  UseMethod("my_new_generic")
}
```

(If you wonder why we have to repeat `my_new_generic` twice, think back to Section \@ref(first-class-functions).)

You don't pass any of the arguments of the generic to `UseMethod()`; it uses deep magic to pass to the method automatically. The precise process is complicated and frequently surprising, so you should avoid doing any computation in a generic. To learn the full details, carefully read the Technical Details section in `?UseMethod`.

### 13.4.1 Method dispatch
\index{S3!method dispatch}
\index{method dispatch!S3}

How does `UseMethod()` work? It basically creates a vector of method names, `paste0("generic", ".", c(class(x), "default"))`, and then looks for each potential method in turn. We can see this in action with `sloop::s3_dispatch()`. You give it a call to an S3 generic, and it lists all the possible methods. For example, what method is called when you print a `Date` object?

```{r}
x <- Sys.Date()
s3_dispatch(print(x))
```

The output here is simple:

* `=>` indicates the method that is called, here `print.Date()`
* `*` indicates a method that is defined, but not called, here `print.default()`.

The "default" class is a special __pseudo-class__. This is not a real class, but is included to make it possible to define a standard fallback that is found whenever a class-specific method is not available.

The essence of method dispatch is quite simple, but as the chapter proceeds you'll see it get progressively more complicated to encompass inheritance, base types, internal generics, and group generics. The code below shows a couple of more complicated cases which we'll come back to in Sections \@ref(inheritance) and \@ref(s3-dispatch). 

```{r}
x <- matrix(1:10, nrow = 2)
s3_dispatch(mean(x))

s3_dispatch(sum(Sys.time()))
```

> s3_dispatch() prints a list of all possible function names that will be considered for method dispatch. There are four possible states:

=> method exists and is found by UseMethod().

-> method exists and is used by NextMethod().

* method exists but is not used.

Nothing (and greyed out in console): method does not exist.

### 13.4.2 Finding methods
\index{S3!methods!locating}

`sloop::s3_dispatch()` lets you find the specific method used for a single call. What if you want to find all methods defined for a generic or associated with a class? That's the job of `sloop::s3_methods_generic()` and `sloop::s3_methods_class()`:

```{r}
s3_methods_generic("mean")

s3_methods_class("ordered")
```

### 13.4.3 Creating methods {#s3-arguments}
\index{S3!methods!creating}
\index{methods!S3}

There are two wrinkles to be aware of when you create a new method:

*   First, you should only ever write a method if you own the generic or the
    class. R will allow you to define a method even if you don't, but it is
    exceedingly bad manners. Instead, work with the author of either the 
    generic or the class to add the method in their code.

*   A method must have the same arguments as its generic. This is enforced in
    packages by `R CMD check`, but it's good practice even if you're not 
    creating a package. 

    There is one exception to this rule: if the generic has `...`, the method 
    can contain a superset of the arguments. This allows methods to take 
    arbitrary additional arguments. The downside of using `...`, however, is 
    that any misspelled arguments will be silently swallowed[^ellipsis],
    as mentioned in Section \@ref(fun-dot-dot-dot).

[^ellipsis]: See <https://github.com/hadley/ellipsis> for an experimental way of warning when methods fail to use all the arguments in `...`, providing a potential resolution of this issue.

### 13.4.4 Exercises

1.  Read the source code for `t()` and `t.test()` and confirm that
    `t.test()` is an S3 generic and not an S3 method. What happens if
    you create an object with class `test` and call `t()` with it? Why?

    ```{r, results = FALSE}
    x <- structure(1:10, class = "test")
    t(x)
    ```

```{r}
t
```

```{r}
t.test
```

```{r}
s3_methods_generic("t.test")
```

```{r}
s3_methods_class("t.test")
```

```{r}
s3_dispatch(t.test(1:10, y = c(7:20)))
```

```{r}
ftype(t.test)
```

```{r}
s3_dispatch(t(x))
```


```{r}
tools::nonS3methods("stats")
```


> `t.test()` is an S3 generic and not an S3 method.

> When we create an object with class test, t() dispatches to the t.default() method. This happens, because UseMethod() simply searches for functions named paste0("generic", ".", c(class(x), "default")).

```{r}
x <- structure(1:10, class = "test")

t(x)
```


2.  What generics does the `table` class have methods for?

```{r}
s3_methods_class("table")
```


3.  What generics does the `ecdf` class have methods for?

```{r}
s3_methods_class("ecdf")
```


4.  Which base generic has the greatest number of defined methods?

```{r}
nrow(s3_methods_generic("plot"))
nrow(s3_methods_generic("print"))
nrow(s3_methods_generic("quantile"))
nrow(s3_methods_generic("summary"))
```

```{r}
base_function <- ls(all.names = TRUE, env = baseenv()) %>% 
  mget(envir = baseenv()) %>% 
  keep(is_function) %>% 
  names()

str(base_function)

s3_generic <- base_function %>% 
  keep(is_s3_generic)

str(s3_generic)
```

```{r}
n_meth_s3_generic <- s3_generic %>% 
  map(~ set_names(nrow(s3_methods_generic(.x)), .x)) %>% 
  flatten_dfr() 

n_meth_s3_generic
str(n_meth_s3_generic)
```


```{r}
long_n_meth_s3_generic <- n_meth_s3_generic %>% 
  pivot_longer(
    everything(), 
    names_to = "generic_name", 
    values_to = "n_methods")

long_n_meth_s3_generic
```


```{r}
long_n_meth_s3_generic %>% 
  arrange(desc(n_methods)) %>% 
  head()
```


```{r}
long_n_meth_s3_generic %>% 
  filter(n_methods == max(n_methods))
```


5.  Carefully read the documentation for `UseMethod()` and explain why the
    following code returns the results that it does. What two usual rules
    of function evaluation does `UseMethod()` violate?

```{r}
g <- function(x,y) {
  x <- 10
  y <- 10
  UseMethod("g",y)
}
g.default <- function(x,y) c(x = x, y = y)
g.factor <- function(x,y) cat("factor")

x <- 1
y <- 1
g(x, factor(c("a","b")))

s3_dispatch(g(x, factor(c("a","b"))))
```

```{r}
g <- function(x) {
  x <- 10
  y <- 10
  UseMethod("g")
}
g.default <- function(x) c(x = x, y = y)

x <- 1
y <- 1
g(x)

s3_dispatch(g(x))
```

```{r}
g <- function(x) {
  x <- 10
  y <- 10
  g2(x)
}

g2 <- function(x) c(x = x, y = y)

x <- 1
y <- 1
g(x)
g2(x)
```

```{r}
g.default(x)
```
> because UseMethod() calls g.default() in a special way so that variables defined inside the generic are available to methods. 

> The exception are arguments supplied to the function: they are passed on as is and cannot be affected by code inside the generic.

`UseMethod()`

> R possesses a simple generic function mechanism which can be used for an object-oriented style of programming. Method dispatch takes place based on the class(es) of the first argument to the generic function or of the object supplied as an argument to UseMethod or NextMethod.

> UseMethod(generic, object)

> An R object is a data object which has a class attribute (and this can be tested by is.object). A class attribute is a character vector giving the names of the classes from which the object inherits. If the object does not have a class attribute, it has an implicit class. Matrices and arrays have class "matrix" or"array" followed by the class of the underlying vector. Most vectors have class the result of mode(x), except that integer vectors have class c("integer", "numeric") and real vectors have class c("double", "numeric").

> When a function calling UseMethod("fun") is applied to an object with class attribute c("first", "second"), the system searches for a function called fun.first and, if it finds it, applies it to the object. If no such function is found a function called fun.second is tried. If no class name produces a suitable function, the function fun.default is used, if it exists, or an error results.

> UseMethod is a primitive function but uses standard argument matching. It is not the only means of dispatch of methods, for there are internal generic and group generic functions. UseMethod currently dispatches on the implicit class even for arguments that are not objects, but the other means of dispatch do not.

```{r}
g <- function(x) {
  x <- 10
  y <- 10
  UseMethod("g")
}

g.factor <- function(x) cat("FACTOR\n")
g.integer <- function(x) cat("INTEGER\n")
g.default <- function(x) cat("NOT THOSE\n")
x <- factor("A")
x2 <- 1L

g(x)
g(x2)
g(list())
g(c())

g.default(x)

```

6.  What are the arguments to `[`? Why is this a hard question to answer?

```{r}
ftype(`[`)
```

```{r}
pryr::show_c_source(.Primitive("["))
```

```{r}
s3_methods_generic("[")
```

```{r}
names(formals(`[.Date`))
names(formals(`[.data.frame`))
```

> the arguments vary with the class of x.

```{r}
visible <- s3_methods_generic("[") %>%
  filter(visible)
visible
```

```{r}
visible_args <- visible %>% 
  mutate(
    method = paste0("[.", class),
    argnames = purrr::map(method, ~ names(formals(.x))),
    args = purrr::map(method, ~ formals(.x)),
    args = purrr::map2(
      argnames, args,
      ~ paste(.x, .y, sep = " = ")
    ),
    args = purrr::set_names(args, method)
  ) 

visible_args
str(visible_args)
```

> pull() is similar to $. It's mostly useful because it looks a little nicer in pipes, it also works with remote data frames, and it can optionally name the output.

```{r}
visible_args %>%
  pull(args) 
```


## 13.5 Object styles
\index{S3!object styles}

So far I've focussed on vector style classes like `Date` and `factor`. These have the key property that `length(x)` represents the number of observations in the vector. There are three variants that do not have this property:

*   Record style objects use a list of equal-length vectors to represent 
    individual components of the object. The best example of this is `POSIXlt`,
    which underneath the hood is a list of 11 date-time components like year, 
    month, and day. Record style classes override `length()` and subsetting 
    methods to conceal this implementation detail.
   
    ```{r}
    x <- as.POSIXlt(ISOdatetime(2020, 1, 1, 0, 0, 1:3))
    x
    
    length(x)
    length(unclass(x))
    
    x[[1]] # the first date time
    unclass(x)[[1]] # the first component, the number of seconds
    ```
    \indexc{POSIXlt}

*   Data frames are similar to record style objects in that both use lists of
    equal length vectors. However, data frames are conceptually two dimensional, 
    and the individual components are readily exposed to the user. The number of
    observations is the number of rows, not the length:
    
    ```{r}
    x <- data.frame(x = 1:100, y = 1:100)
    length(x)
    nrow(x)
    ```
    \indexc{Date}

*   Scalar objects typically use a list to represent a single thing. 
    For example, an `lm` object is a list of length 12 but it represents one
    model.
    
    ```{r}
    mod <- lm(mpg ~ wt, data = mtcars)
    length(mod)
    ```
    
    Scalar objects can also be built on top of functions, calls, and 
    environments[^s3-pairlist]. This is less generally useful, but you can see 
    applications in `stats::ecdf()`, R6 (Chapter \@ref(r6)), and
    `rlang::quo()` (Chapter \@ref(quasiquotation)).
    \indexc{lm()}
    
[^s3-pairlist]: You can also build an object on top of a pairlist, but I have yet to find a good reason to do so.

Unfortunately, describing the appropriate use of each of these object styles is beyond the scope of this book. However, you can learn more from the documentation of the vctrs package (<https://vctrs.r-lib.org>); the package also provides constructors and helpers that make implementation of the different styles easier.

### 13.5.1 Exercises

1.  Categorise the objects returned by `lm()`, `factor()`, `table()`, 
    `as.Date()`, `as.POSIXct()` `ecdf()`, `ordered()`, `I()` into the
    styles described above.

* Vector style objects: `factor()`, `table()`, `as.Date()`,  `ordered()`

> For vector style classes, length(x) represents the number of observations. 

```{r}
factor <- factor(letters[1:20])
str(factor)
length(factor)
```

```{r}
table <- table(letters[1:20])
str(table)
length(table)
```

```{r}
x <- c("1jan1960", "2jan1960", "31mar1960", "30jul1960")
z <- as.Date(x, "%d%b%Y")

str(z)
length(z)
length(unclass(z))
```



```{r}
ordered <- ordered(rnorm(12))
str(ordered)
length(ordered)

```


* Record style objects: `as.POSIXct()` 

> Record style objects use a list of equal length elements to represent individual components.

```{r}
now <- as.POSIXlt(z)

str(now)
length(now)
length(unclass(now))
```

* Data frame objects: NA

> For data frames and matrices, the observations are represented by the rows.

* Scalar objects: `lm()`, `ecdf()`

> Scalar style objects use a list to represent a single thing.

```{r}
mod <- lm(mpg ~ wt, data = mtcars)
str(mod)
length(mod)
```

```{r}
x <- rnorm(12)
Fn <- ecdf(x)
str(Fn)
length(Fn)
```

> `I(x)` depends on the input

> Change the class of an object to indicate that it should be treated ‘as is’.

2.  What would a constructor function for `lm` objects, `new_lm()`, look like?
    Use `?lm` and experimentation to figure out the required fields and their
    types.
    
```{r}
lm
```

```{r}
mod <- lm(mpg ~ wt, data = mtcars)
typeof(mod)
length(mod)
```
```{r}
summary(mod)
```

```{r}
str(mod)
```


```{r}
attributes(mod)
```



```{r}
new_lm <- function(coefficients, 
                   residuals, 
                   effects, 
                   rank, 
                   fitted.values, 
                   assign,
                   qr, 
                   df.residual, 
                   xlevels, 
                   call, 
                   terms, 
                   model) {
  
  stopifnot(is.double(coefficients), 
            is.double(residuals), 
            is.double(effects), 
            is.integer(rank), 
            is.double(fitted.values),
            is.integer(assign), 
            is.list(qr), 
            is.integer(df.residual),
            is.list(xlevels), 
            is.language(call), 
            is.language(terms),
            is.list(model)
  )
  
  structure(
    list(
      coefficients = coefficients,
      residuals = residuals,
      effects = effects,
      rank = rank, 
      fitted.values = fitted.values,
      assign = assign,
      qr = qr,
      df.residual = df.residual,
      xlevels = xlevels,
      call = call,
      terms = terms, 
      model = model
    ),
    class = "lm"
  )
}
```

