---
title: "Ch6_Functions_2"
author: "Min-Yao"
date: "2022-10-16"
output: html_document
---

## `...`  (dot-dot-dot) {#fun-dot-dot-dot}
\indexc{...}
\index{functions!variadic|see {...}}
\index{ellipsis|see {...}}
\index{dot-dot-dot|see {...}}

Functions can have a special argument `...` (pronounced dot-dot-dot). With it, a function can take any number of additional arguments. In other programming languages, this type of argument is often called _varargs_ (short for variable arguments), and a function that uses it is said to be variadic. 

You can also use `...` to pass those additional arguments on to another function.

```{r}
i01 <- function(y, z) {
  list(y = y, z = z)
}

i02 <- function(x, ...) {
  i01(...)
}

str(i02(x = 1, y = 2, z = 3))
```

Using a special form, `..N`, it's possible (but rarely useful) to refer to elements of `...` by position:

```{r}
i03 <- function(...) {
  list(first = ..1, third = ..3)
}
str(i03(1, 2, 3))
```

More useful is `list(...)`, which evaluates the arguments and stores them in a list:

```{r}
i04 <- function(...) {
  list(...)
}
str(i04(a = 1, b = 2))
```

(See also `rlang::list2()` to support splicing and to silently ignore trailing commas, and `rlang::enquos()` to capture unevaluated arguments, the topic of [quasiquotation].)

There are two primary uses of `...`, both of which we'll come back to later in the book:

*   If your function takes a function as an argument, you want some way to 
    pass additional arguments to that function. In this example, `lapply()`
    uses `...` to pass `na.rm` on to `mean()`:
    
    ```{r}
    x <- list(c(1, 3, NA), c(4, NA, 6))
    str(lapply(x, mean, na.rm = TRUE))
    ```
    
    We'll come back to this technique in Section \@ref(passing-arguments).
    
*   If your function is an S3 generic, you need some way to allow methods to 
    take arbitrary extra arguments. For example, take the `print()` function. 
    Because there are different options for printing depending on the type of 
    object, there's no way to pre-specify every possible argument and `...` 
    allows individual methods to have different arguments:

    ```{r, eval = FALSE}
    print(factor(letters), max.levels = 4)
    
    print(y ~ x, showEnv = TRUE)
    ```
    
    We'll come back to this use of `...` in Section \@ref(s3-arguments).

Using `...` comes with two downsides:

*   When you use it to pass arguments to another function, you have to 
    carefully explain to the user where those arguments go. This makes it
    hard to understand what you can do with functions like `lapply()` and 
    `plot()`.
    
*   A misspelled argument will not raise an error. This makes it easy for 
    typos to go unnoticed:

    ```{r}
    sum(1, 2, NA, na_rm = TRUE)
    ```

### Exercises

1.  Explain the following results:
    
    ```{r}
    sum(1, 2, 3)
    mean(1, 2, 3)
    
    sum(1, 2, 3, na.omit = TRUE)
    mean(1, 2, 3, na.omit = TRUE)
    ```

1.  Explain how to find the documentation for the named arguments in the 
    following function call:
    
    ```{r, fig.asp = 1, small_mar = TRUE, fig.width = 3}
    plot(1:10, col = "red", pch = 20, xlab = "x", col.lab = "blue")
    ```
    
1.  Why does `plot(1:10, col = "red")` only colour the points, not the axes 
    or labels? Read the source code of `plot.default()` to find out.

## Exiting a function

Most functions exit in one of two ways[^esoterica]: they either return a value, indicating success, or they throw an error, indicating failure. This section describes return values (implicit versus explicit; visible versus invisible), briefly discusses errors, and introduces exit handlers, which allow you to run code when a function exits.

[^esoterica]: Functions can exit in other more esoteric ways like signalling a condition that is caught by an exit handler, invoking a restart, or pressing "Q" in an interactive browser.

### Implicit versus explicit returns
\index{functions!return value}
\indexc{return()}

There are two ways that a function can return a value:

*   Implicitly, where the last evaluated expression is the return value:

    ```{r}
    j01 <- function(x) {
      if (x < 10) {
        0
      } else {
        10
      }
    }
    j01(5)
    j01(15)
    ```

*   Explicitly, by calling `return()`:

    ```{r}
    j02 <- function(x) {
      if (x < 10) {
        return(0)
      } else {
        return(10)
      }
    }
    ```
    
### Invisible values {#invisible}
\indexc{invisible()} 
\index{functions!invisible results}
\index{assignment}

Most functions return visibly: calling the function in an interactive context prints the result.

```{r}
j03 <- function() 1
j03()
```

However, you can prevent automatic printing by applying `invisible()` to the last value:

```{r}
j04 <- function() invisible(1)
j04()
```

To verify that this value does indeed exist, you can explicitly print it or wrap it in parentheses:

```{r}
print(j04())

(j04())
```

Alternatively, you can use `withVisible()` to return the value and a visibility flag:

```{r}
str(withVisible(j04()))
```

The most common function that returns invisibly is `<-`: 

```{r}
a <- 2
(a <- 2)
```

This is what makes it possible to chain assignments:

```{r}
a <- b <- c <- d <- 2
```

In general, any function called primarily for a side effect (like `<-`, `print()`, or `plot()`) should return an invisible value (typically the value of the first argument).

### Errors
\indexc{stop()}
\index{errors}

If a function cannot complete its assigned task, it should throw an error with `stop()`, which immediately terminates the execution of the function.

```{r, error = TRUE}
j05 <- function() {
  stop("I'm an error")
  return(10)
}
j05()
```

An error indicates that something has gone wrong, and forces the user to deal with the problem. Some languages (like C, Go, and Rust) rely on special return values to indicate problems, but in R you should always throw an error. You'll learn more about errors, and how to handle them, in Chapter \@ref(conditions).

### Exit handlers {#on-exit}
\indexc{on.exit()}
\index{handlers!exit}

Sometimes a function needs to make temporary changes to the global state. But having to cleanup those changes can be painful (what happens if there's an error?). To ensure that these changes are undone and that the global state is restored no matter how a function exits, use `on.exit()` to set up an __exit handler__. The following simple example shows that the exit handler is run regardless of whether the function exits normally or with an error.

```{r, error = TRUE}
j06 <- function(x) {
  cat("Hello\n")
  on.exit(cat("Goodbye!\n"), add = TRUE)
  
  if (x) {
    return(10)
  } else {
    stop("Error")
  }
}

j06(TRUE)

j06(FALSE)
```

::: sidebar
Always set `add = TRUE` when using `on.exit()`. If you don't, each call to `on.exit()` will overwrite the previous exit handler. Even when only registering a single handler, it's good practice to set `add = TRUE` so that you won't get any unpleasant surprises if you later add more exit handlers.
:::

`on.exit()` is useful because it allows you to place clean-up code directly next to the code that requires clean-up:

```{r}
cleanup <- function(dir, code) {
  old_dir <- setwd(dir)
  on.exit(setwd(old_dir), add = TRUE)
  
  old_opt <- options(stringsAsFactors = FALSE)
  on.exit(options(old_opt), add = TRUE)
}
```

Coupled with lazy evaluation, this creates a very useful pattern for running a block of code in an altered environment:

```{r}
with_dir <- function(dir, code) {
  old <- setwd(dir)
  on.exit(setwd(old), add = TRUE)

  force(code)
}

getwd()
with_dir("~", getwd())
```

The use of `force()` isn't strictly necessary here as simply referring to `code` will force its evaluation. However, using `force()` makes it very clear that we are deliberately forcing the execution. You'll learn other uses of `force()` in Chapter \@ref(function-factories).

The withr package [@withr] provides a collection of other functions for setting up a temporary state.

In R 3.4 and earlier, `on.exit()` expressions are always run in order of creation:

```{r}
j08 <- function() {
  on.exit(message("a"), add = TRUE)
  on.exit(message("b"), add = TRUE)
}
j08()
```

This can make cleanup a little tricky if some actions need to happen in a specific order; typically you want the most recent added expression to be run first. In R 3.5 and later, you can control this by setting `after = FALSE`:

```{r, eval = getRversion() >= "3.5"}
j09 <- function() {
  on.exit(message("a"), add = TRUE, after = FALSE)
  on.exit(message("b"), add = TRUE, after = FALSE)
}
j09()
```

### Exercises

1.  What does `load()` return? Why don't you normally see these values?

1.  What does `write.table()` return? What would be more useful?

1.  How does the `chdir` parameter of `source()` compare to `with_dir()`? Why 
    might you prefer one to the other?

1.  Write a function that opens a graphics device, runs the supplied code, and 
    closes the graphics device (always, regardless of whether or not the 
    plotting code works).

1.  We can use `on.exit()` to implement a simple version of `capture.output()`.

    ```{r, eval = getRversion() >= "3.5"}
    capture.output2 <- function(code) {
      temp <- tempfile()
      on.exit(file.remove(temp), add = TRUE, after = TRUE)

      sink(temp)
      on.exit(sink(), add = TRUE, after = TRUE)

      force(code)
      readLines(temp)
    }
    capture.output2(cat("a", "b", "c", sep = "\n"))
    ```

    Compare `capture.output()` to `capture.output2()`. How do the functions 
    differ? What features have I removed to make the key ideas easier to see? 
    How have I rewritten the key ideas so they're easier to understand?
    
## Function forms

> To understand computations in R, two slogans are helpful:
>
> * Everything that exists is an object.
> * Everything that happens is a function call.
>
> --- John Chambers

While everything that happens in R is a result of a function call, not all calls look the same. Function calls come in four varieties:

* __prefix__: the function name comes before its arguments, like
  `foofy(a, b, c)`. These constitute of the majority of function calls in R.

* __infix__: the function name comes in between its arguments, like
  `x + y`. Infix forms are used for many mathematical operators, and for
  user-defined functions that begin and end with `%`.

* __replacement__: functions that replace values by assignment, like
  `names(df) <- c("a", "b", "c")`. They actually look like prefix functions.

* __special__: functions like `[[`, `if`, and `for`. While they don't have a
  consistent structure, they play important roles in R's syntax.

While there are four forms, you actually only need one because any call can be written in prefix form. I'll demonstrate this property, and then you'll learn about each of the forms in turn.

### Rewriting to prefix form {#prefix-transform}
\index{'@\texttt{`}}
\index{backticks|see {\texttt{`}}}
\indexc{sapply()}

An interesting property of R is that every infix, replacement, or special form can be rewritten in prefix form. Doing so is useful because it helps you better understand the structure of the language, it gives you the real name of every function, and it allows you to modify those functions for fun and profit.

The following example shows three pairs of equivalent calls, rewriting an infix form, replacement form, and a special form into prefix form. 

```{r, eval = FALSE}
x + y
`+`(x, y)

names(df) <- c("x", "y", "z")
`names<-`(df, c("x", "y", "z"))

for(i in 1:10) print(i)
`for`(i, 1:10, print(i))
```

Suprisingly, in R, `for` can be called like a regular function! The same is true for basically every operation in R, which means that knowing the function name of a non-prefix function allows you to override its behaviour. For example, if you're ever feeling particularly evil, run the following code while a friend is away from their computer. It will introduce a fun bug: 10% of the time, it will add 1 to any numeric calculation inside the parentheses.

```{r}
`(` <- function(e1) {
  if (is.numeric(e1) && runif(1) < 0.1) {
    e1 + 1
  } else {
    e1
  }
}
replicate(50, (1 + 2))
rm("(")
```

Of course, overriding built-in functions like this is a bad idea, but, as you'll learn in Section \@ref(html-env), it's possible to apply it only to selected code blocks. This provides a clean and elegant approach to writing domain specific languages and translators to other languages.

A more useful application comes up when using functional programming tools. For example, you could use `lapply()` to add 3 to every element of a list by first defining a function `add()`:

```{r}
add <- function(x, y) x + y
lapply(list(1:3, 4:5), add, 3)
```

But we can also get the same result simply by relying on the existing `+` function:

```{r}
lapply(list(1:3, 4:5), `+`, 3)
```

We'll explore this idea in detail in Section \@ref(functionals).

### Prefix form {#prefix-form}
\index{functions!arguments}
\index{arguments!matching}

The prefix form is the most common form in R code, and indeed in the majority of programming languages. Prefix calls in R are a little special because you can specify arguments in three ways:

* By position, like `help(mean)`.
* Using partial matching, like `help(top = mean)`.
* By name, like `help(topic = mean)`.

As illustrated by the following chunk, arguments are matched by exact name, then with unique prefixes, and finally by position.

```{r, error = TRUE}
k01 <- function(abcdef, bcde1, bcde2) {
  list(a = abcdef, b1 = bcde1, b2 = bcde2)
}
str(k01(1, 2, 3))
str(k01(2, 3, abcdef = 1))

# Can abbreviate long argument names:
str(k01(2, 3, a = 1))

# But this doesn't work because abbreviation is ambiguous
str(k01(1, 3, b = 1))
```

In general, use positional matching only for the first one or two arguments; they will be the most commonly used, and most readers will know what they are. Avoid using positional matching for less commonly used arguments, and never use partial matching. Unfortunately you can't disable partial matching, but you can turn it into a warning with the `warnPartialMatchArgs` option:
\index{options!warnPartialMatchArgs@\texttt{warnPartialMatchArgs}}

```{r}
options(warnPartialMatchArgs = TRUE)
x <- k01(a = 1, 2, 3)
```

### Infix functions
\index{functions!infix} 
\index{infix functions} 
\indexc{\%\%}

Infix functions get their name from the fact the function name comes inbetween its arguments, and hence have two arguments. R comes with a number of built-in infix operators: `:`, `::`, `:::`, `$`, `@`, `^`, `*`, `/`, `+`, `-`, `>`, `>=`, `<`, `<=`, `==`, `!=`, `!`, `&`, `&&`, `|`, `||`, `~`, `<-`, and `<<-`. You can also create your own infix functions that start and end with `%`. Base R uses this pattern to define `%%`, `%*%`, `%/%`, `%in%`, `%o%`, and `%x%`.

Defining your own infix function is simple. You create a two argument function and bind it to a name that starts and ends with `%`:

```{r}
`%+%` <- function(a, b) paste0(a, b)
"new " %+% "string"
```

The names of infix functions are more flexible than regular R functions: they can contain any sequence of characters except for `%`. You will need to escape any special characters in the string used to define the function, but not when you call it:

```{r}
`% %` <- function(a, b) paste(a, b)
`%/\\%` <- function(a, b) paste(a, b)

"a" % % "b"
"a" %/\% "b"
```

R's default precedence rules mean that infix operators are composed left to right:

```{r}
`%-%` <- function(a, b) paste0("(", a, " %-% ", b, ")")
"a" %-% "b" %-% "c"
```

There are two special infix functions that can be called with a single argument: `+` and `-`.  
```{r}
-1
+10
```

### Replacement functions {#replacement-functions}
\index{replacement functions} 
\index{functions!replacement}
\index{assignment!in replacement functions}

Replacement functions act like they modify their arguments in place, and have the special name `xxx<-`. They must have arguments named `x` and `value`, and must return the modified object. For example, the following function modifies the second element of a vector: 

```{r}
`second<-` <- function(x, value) {
  x[2] <- value
  x
}
```

Replacement functions are used by placing the function call on the left side of `<-`: 

```{r}
x <- 1:10
second(x) <- 5L
x
```

I say they act like they modify their arguments in place, because, as explained in Section \@ref(modify-in-place), they actually create a modified copy. We can see that by using `tracemem()`:

```{r, eval = FALSE}
x <- 1:10
tracemem(x)
#> <0x7ffae71bd880>

second(x) <- 6L
#> tracemem[0x7ffae71bd880 -> 0x7ffae61b5480]: 
#> tracemem[0x7ffae61b5480 -> 0x7ffae73f0408]: second<- 
```

If your replacement function needs additional arguments, place them between `x` and `value`, and call the replacement function with additional arguments on the left:

```{r}
`modify<-` <- function(x, position, value) {
  x[position] <- value
  x
}
modify(x, 1) <- 10
x
```

When you write `modify(x, 1) <- 10`, behind the scenes R turns it into:

```{r, eval = FALSE}
x <- `modify<-`(x, 1, 10)
```

Combining replacement with other functions requires more complex translation. For example:

```{r}
x <- c(a = 1, b = 2, c = 3)
names(x)

names(x)[2] <- "two"
names(x)
```

is translated into:

```{r}
`*tmp*` <- x
x <- `names<-`(`*tmp*`, `[<-`(names(`*tmp*`), 2, "two"))
rm(`*tmp*`)
```

(Yes, it really does create a local variable named `*tmp*`, which is removed afterwards.)

### Special forms
\index{special forms}
\index{functions!special}

Finally, there are a bunch of language features that are usually written in special ways, but also have prefix forms. These include parentheses:

* `(x)` (`` `(`(x) ``)
* `{x}` (`` `{`(x) ``).

The subsetting operators:

* `x[i]` (`` `[`(x, i) ``) 
* `x[[i]]` (`` `[[`(x, i) ``)

And the tools of control flow:

* `if (cond) true` (`` `if`(cond, true) ``)
* `if (cond) true else false` (`` `if`(cond, true, false) ``)
* `for(var in seq) action` (`` `for`(var, seq, action) ``)
* `while(cond) action` (`` `while`(cond, action)  ``)
* `repeat expr` (`` `repeat`(expr) ``)
* `next` (`` `next`() ``)
* `break` (`` `break`() ``)

Finally, the most complex is the `function` function:

* `function(arg1, arg2) {body}` (`` `function`(alist(arg1, arg2), body, env)``)

Knowing the name of the function that underlies a special form is useful for getting documentation: `?(` is a syntax error; `` ?`(` `` will give you the documentation for parentheses.

All special forms are implemented as primitive functions (i.e. in C); this means printing these functions is not informative:

```{r}
`for`
```

### Exercises

1. Rewrite the following code snippets into prefix form:

    ```{r, eval = FALSE}
    1 + 2 + 3
    
    1 + (2 + 3)
    
    if (length(x) <= 5) x[[5]] else x[[n]]
    ```

1.  Clarify the following list of odd function calls:

    ```{r, eval = FALSE}
    x <- sample(replace = TRUE, 20, x = c(1:10, NA))
    y <- runif(min = 0, max = 1, 20)
    cor(m = "k", y = y, u = "p", x = x)
    ```


1. Explain why the following code fails:

    ```{r, eval = FALSE}
    modify(get("x"), 1) <- 10
    #> Error: target of assignment expands to non-language object
    ```

1. Create a replacement function that modifies a random location in a vector.

1. Write your own version of `+` that pastes its inputs together if they are 
   character vectors but behaves as usual otherwise. In other words, make this 
   code work:
   
    ```{r, eval = FALSE}
    1 + 2
    #> [1] 3
    
    "a" + "b"
    #> [1] "ab"
    ```

1. Create a list of all the replacement functions found in the base package. 
   Which ones are primitive functions? (Hint: use `apropos()`.)

1. What are valid names for user-created infix functions?

1. Create an infix `xor()` operator.

1. Create infix versions of the set functions `intersect()`, `union()`, and
   `setdiff()`. You might call them `%n%`, `%u%`, and `%/%` to match 
   conventions from mathematics.


